---
title: "Session 11 — Fourier seasonality (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

## Parameter counts: dummies vs Fourier

```{r}
param_counts <- tibble::tibble(
  m = c(4, 52),
  label = c("quarterly", "weekly"),
  dummies = m - 1,
  fourier_K1 = 2 * 1,
  fourier_K2 = 2 * 2
)
param_counts
```

Expect `dummies = 3, 51` and `fourier_K2 = 4, 4` for $m=4,52$ — a $\sim92\%$ reduction at
$m=52$, $K=2$.

## The Nyquist coincidence at $K = m/2$ (quarterly, $m=4$)

```{r}
m <- 4
K <- m / 2
effective_params <- 2 * K - 1
c(m = m, K = K, effective_params = effective_params, dummy_params = m - 1)
```

`effective_params` should equal `dummy_params` (both `3`): at the top harmonic, the sine
column is identically zero for integer $t$, so it drops out automatically.

# Data: Australian beer production

```{r}
beer <- aus_production |> filter(!is.na(Beer))
beer |> autoplot(Beer) + labs(title = "Australian beer production")
```

# Season vs Fourier K = 1, 2

```{r}
fit <- beer |>
  model(
    dummies = TSLM(Beer ~ trend() + season()),
    f1 = TSLM(Beer ~ trend() + fourier(K = 1)),
    f2 = TSLM(Beer ~ trend() + fourier(K = 2))
  )
glance(fit) |> select(.model, AICc) |> arrange(AICc)
```

Confirm `dummies` and `f2` have (numerically) the same AICc — same fitted values in two
different bases, as predicted by the Nyquist argument above.

```{r}
fitted_wide <- fit |> select(dummies, f2) |>
  augment() |>
  as_tibble() |>
  select(.model, Quarter, .fitted) |>
  tidyr::pivot_wider(names_from = .model, values_from = .fitted)
max(abs(fitted_wide$dummies - fitted_wide$f2))
```

This maximum absolute difference should be tiny (numerical noise only).

```{r}
fit |> forecast(h = "3 years") |> autoplot(beer, level = NULL) +
  labs(title = "Dummies vs Fourier K = 1, 2")
```

# Residual seasonality check

```{r}
fit |> select(f1) |> gg_tsresiduals()
```

# Weekly-style discussion

For $m=52$, $K=2$ needs 4 parameters versus 51 dummy columns. Dynamic harmonic regression
(Session 12) combines a small-$K$ Fourier seasonal term with ARIMA errors for the
remaining short-run dynamics — this is the standard approach for weekly series with
annual seasonality.

# Your turn

1. Compute the parameter-count table for $m=7$ (daily-of-week) and $m=365$ (daily-of-year).
2. For Beer, does `f1` show more residual seasonality than `f2` in `gg_tsresiduals()`? Explain using the
   parameter-count argument.
3. At what $K$ would you stop increasing harmonics for a weekly series, and why (hint: AICc plateau)?
