---
title: "Session 5 — Seasonal ARIMA (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.  
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

## Combined seasonal + regular difference (Beer 1993)

```{r}
beer_toy <- aus_production |>
  filter(Quarter >= yearquarter("1992 Q1"),
         Quarter <= yearquarter("1993 Q4")) |>
  select(Quarter, Beer)

beer_toy |>
  mutate(
    d4 = difference(Beer, lag = 4),
    d14 = difference(difference(Beer, lag = 4))
  )
```

Check 1993 Q2: $(1-B)(1-B^4)y_t = 21$; Q3: $-21$.

## Airline MA product (toy)

Expand $(1+0.4B)(1+0.6B^4)$ applied to white noise:

```{r}
eps <- c(1, -1, 0.5, 0, 2)
theta <- 0.4
Theta <- 0.6
# z_t = e_t + theta e_{t-1} + Theta e_{t-4} + theta*Theta e_{t-5}
z <- eps +
  theta * dplyr::lag(eps, 1, default = 0) +
  Theta * dplyr::lag(eps, 4, default = 0) +
  theta * Theta * dplyr::lag(eps, 5, default = 0)
tibble::tibble(t = 1:5, eps = eps, z = z)
```

Expected $z_5 = 2.6$.

# Australian beer bake-off

```{r}
beer <- aus_production |> filter(!is.na(Beer))
beer |> gg_tsdisplay(Beer, plot_type = "partial")

train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)

fit <- train |>
  model(
    snaive = SNAIVE(Beer),
    ets = ETS(Beer),
    sarima = ARIMA(Beer),
    airline = ARIMA(Beer ~ pdq(0, 1, 1) + PDQ(0, 1, 1))
  )

glance(fit) |> arrange(AICc)
fc <- fit |> forecast(h = nrow(test))
accuracy(fc, beer) |> arrange(RMSE)
fc |> autoplot(beer, level = NULL) +
  labs(title = "Beer: ETS vs SARIMA vs SNAIVE vs airline")
```

```{r}
fit |> select(sarima) |> gg_tsresiduals()
report(fit |> select(airline))
```

# Your turn

1. Try another seasonal series (`us_employment` industry, or tourism).
2. Fix PDQ manually and compare to `ARIMA()` auto.
3. Draft 5 bullets for Assignment A1.
