---
title: "Session 5 — Seasonal ARIMA (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.  
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

## Combined seasonal + regular difference (Beer 1993)

```{r}
beer_toy <- aus_production |>
  filter(Quarter >= yearquarter("1992 Q1"),
         Quarter <= yearquarter("1993 Q4")) |>
  select(Quarter, Beer)

beer_toy |>
  mutate(
    d4 = difference(Beer, lag = 4),
    d14 = difference(difference(Beer, lag = 4))
  )
```

Check 1993 Q2: $(1-B)(1-B^4)y_t = 21$; Q3: $-21$.

## Airline MA product (toy)

Expand $(1+0.4B)(1+0.6B^4)$ applied to white noise:

```{r}
eps <- c(1, -1, 0.5, 0, 2)
theta <- 0.4
Theta <- 0.6
# z_t = e_t + theta e_{t-1} + Theta e_{t-4} + theta*Theta e_{t-5}
z <- eps +
  theta * dplyr::lag(eps, 1, default = 0) +
  Theta * dplyr::lag(eps, 4, default = 0) +
  theta * Theta * dplyr::lag(eps, 5, default = 0)
tibble::tibble(t = 1:5, eps = eps, z = z)
```

Expected $z_5 = 2.6$.

# Australian beer bake-off

```{r}
beer <- aus_production |> filter(!is.na(Beer))
beer |> gg_tsdisplay(Beer, plot_type = "partial")

train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)

fit <- train |>
  model(
    snaive = SNAIVE(Beer),
    ets = ETS(Beer),
    sarima = ARIMA(Beer),
    airline = ARIMA(Beer ~ pdq(0, 1, 1) + PDQ(0, 1, 1))
  )

glance(fit) |> arrange(AICc)
fc <- fit |> forecast(h = nrow(test))
accuracy(fc, beer) |> arrange(RMSE)
fc |> autoplot(beer, level = NULL) +
  labs(title = "Beer: ETS vs SARIMA vs SNAIVE vs airline")
```

```{r}
# Posit Cloud: augment() + feasts::ACF() gets an empty series for this SARIMA.
# Use stored innovations when present; otherwise refit the same orders with stats::arima.
m <- fit[["sarima"]][[1]]
e <- numeric(0)
est <- tryCatch(m$fit$est, error = function(err) NULL)
if (!is.null(est) && ".resid" %in% names(est)) e <- as.numeric(est$.resid)
e <- e[is.finite(e)]

if (length(e) < 2) {
  y <- train$Beer
  y <- y[is.finite(y)]
  sp <- tryCatch(as.data.frame(m$fit$spec), error = function(err) NULL)
  p <- if (!is.null(sp) && "p" %in% names(sp)) sp$p[[1]] else 0
  d <- if (!is.null(sp) && "d" %in% names(sp)) sp$d[[1]] else 1
  q <- if (!is.null(sp) && "q" %in% names(sp)) sp$q[[1]] else 1
  P <- if (!is.null(sp) && "P" %in% names(sp)) sp$P[[1]] else 0
  D <- if (!is.null(sp) && "D" %in% names(sp)) sp$D[[1]] else 1
  Q <- if (!is.null(sp) && "Q" %in% names(sp)) sp$Q[[1]] else 1
  per <- if (!is.null(sp) && "period" %in% names(sp)) sp$period[[1]] else 4
  fit_s <- stats::arima(
    y,
    order = c(p, d, q),
    seasonal = list(order = c(P, D, Q), period = per)
  )
  e <- as.numeric(stats::residuals(fit_s))
  e <- e[is.finite(e)]
}

qtr <- tail(train$Quarter, length(e))
innov <- tibble::tibble(Quarter = qtr, e = e)

ggplot(innov, aes(Quarter, e)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line() +
  labs(title = "Auto SARIMA — innovation residuals", y = NULL)

acf_obj <- stats::acf(innov$e, plot = FALSE)
acf_df <- tibble::tibble(
  lag = as.numeric(acf_obj$lag),
  acf = as.numeric(acf_obj$acf)
)
acf_df <- acf_df[acf_df$lag > 0, ]
band <- 2 / sqrt(length(innov$e))
ggplot(acf_df, aes(lag, acf)) +
  geom_hline(yintercept = c(-band, 0, band), linetype = c(2, 1, 2), colour = "steelblue") +
  geom_col(width = 0.25) +
  labs(title = "ACF of innovations", x = "lag", y = "acf")

report(fit |> select(airline))
```

# Your turn

1. Try another seasonal series (`us_employment` industry, or tourism).
2. Fix PDQ manually and compare to `ARIMA()` auto.
3. Draft 5 bullets for Assignment A1.
