---
title: "Lab 05 — Seasonal ARIMA"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(fpp3)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4
)
source("lab_submission.R")

lab_meta <- list(id = "lab05", title = "Seasonal ARIMA")
lab_prompts <- c(
  q1 = "The quarterly airline model (0,1,1)(0,1,1)_4 is:",
  q2 = "Beer 1993 Q2: 421 - 433 - 410 + 443 equals:",
  q3 = "After differencing, an ACF spike at lag m often suggests:",
  q4 = "Which Beer model would you ship, and what evidence (hold-out or residuals) would you cite?"
)
```

## What you will do

Expand the airline operator and the Beer seasonal difference. Residual plots stay inside exercise chunks.

Open this file and click **Run Document**. Packages: `fpp3`, `learnr`.
Exercise chunks run when you click Run Code. They are not knitted in one pass.

## Seasonal difference

```{r beer-diff, exercise=TRUE, exercise.lines=10}
beer <- aus_production |>
  filter(!is.na(Beer)) |>
  select(Quarter, Beer) |>
  mutate(d14 = difference(difference(Beer, lag = 4)))
beer |> filter_index("1993 Q1" ~ "1993 Q2")
```

```{r q1}
question(
  lab_prompts[["q1"]],
  answer("(1-B)(1-B^4) y_t = (1+theta B)(1+Theta B^4) e_t", correct = TRUE),
  answer("(1-B) y_t = (1+theta B) e_t"),
  answer("y_t = phi y_{t-4} + e_t"),
  answer("ETS(A,N,N)"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
```{r q2}
question(
  lab_prompts[["q2"]],
  answer("21", correct = TRUE),
  answer("-21"),
  answer("443"),
  answer("0"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
```{r q3}
question(
  lab_prompts[["q3"]],
  answer("Seasonal MA, often Q=1", correct = TRUE),
  answer("Set nonseasonal p = m"),
  answer("Set d=2"),
  answer("Drop the seasonal period"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
```{r q4}
question_text(
  lab_prompts[["q4"]],
  open_answer(),
  placeholder = "SNAIVE vs seasonal ETS vs SARIMA on the same hold-out.",
  trim = TRUE
)
```

## Submit

Answer the questions above first, type your name, download the `.txt`, and upload it.

```{r submit-ui, echo=FALSE}
shiny::textInput("student_name", "Your name", placeholder = "Name Surname")
shiny::downloadButton("download_txt", "Download submission .txt")
```

```{r submit-server, context="server"}
output$download_txt <- shiny::downloadHandler(
  filename = function() lab_submission_filename(input$student_name, lab_meta$id),
  content = function(file) {
    write_lab_submission(
      path = file,
      name = input$student_name,
      lab_id = lab_meta$id,
      lab_title = lab_meta$title,
      state = learnr::get_tutorial_state(),
      prompts = lab_prompts
    )
  }
)
```
