---
title: "Session 14 - Final review lab"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Slides summarise the course checklist. This lab is a reproducible
> capstone bake-off using the same Beer hold-out as Sessions 5 and 13.

# Packages

```{r}
library(fpp3)
```

# Capstone bake-off on Beer (train through 2007 Q4)

```{r}
beer <- aus_production |> filter(!is.na(Beer))
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)
h <- nrow(test)

fit <- train |>
  model(
    snaive = SNAIVE(Beer),
    ets = ETS(Beer),
    arima = ARIMA(Beer),
    dhr = ARIMA(Beer ~ fourier(K = 2) + PDQ(0, 0, 0)),
    tslm = TSLM(Beer ~ trend() + season())
  )
fc <- fit |> forecast(h = h)
accuracy(fc, beer) |> select(.model, RMSE, MAE, MAPE) |> arrange(RMSE)
fc |> autoplot(beer, level = NULL) +
  labs(title = "Capstone bake-off on Beer (hold-out from 2008 Q1)")
```

# Residual check (lowest-RMSE model)

```{r}
best <- accuracy(fc, beer) |> arrange(RMSE) |> slice(1) |> pull(.model)
fit |> select(all_of(best)) |> gg_tsresiduals()
```

# Hand reminder: combined difference (Session 2)

For Beer 1993 Q2, verify $(1-B)(1-B^4)y_t = 21$:

```{r}
y <- aus_production |>
  filter(Quarter >= yearquarter("1992 Q1"),
         Quarter <= yearquarter("1993 Q2")) |>
  pull(Beer)
n <- length(y)
y[n] - y[n - 1] - y[n - 4] + y[n - 5]
```

# Optional hierarchical glimpse (tourism)

```{r}
tourism_agg <- tourism |>
  aggregate_key(Purpose, Trips = sum(Trips))
tourism_agg |>
  filter(is_aggregated(Purpose)) |>
  autoplot(Trips) +
  labs(title = "Australian tourism trips (all purposes)")
```

# Optional VAR glimpse

```{r}
fit_var <- us_change |>
  model(var = VAR(vars(Consumption, Income)))
fit_var |> forecast(h = 8) |>
  autoplot(us_change, level = NULL) +
  labs(title = "VAR(Consumption, Income): 8-quarter forecast")
```

# Your turn

1. Which model wins on RMSE? Would you still recommend it after looking at residuals?
2. Name one case where TSLM might beat SARIMA on interpretability even if RMSE is worse.
3. Write a one-page recommendation memo for Beer forecasts (method + why + risks).
