---
title: "Session 1 — ETS and AIC (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.  
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
# install.packages("fpp3")  # once
library(fpp3)
```

# Hand calculations (match the slides)

## SES toy series

Data: $y = (10, 12, 11, 13)$, $\alpha = 0.5$, $\ell_0 = 10$.

```{r}
y <- c(10, 12, 11, 13)
alpha <- 0.5
ell <- numeric(4)
ell0 <- 10
ell[1] <- alpha * y[1] + (1 - alpha) * ell0
for (t in 2:4) {
  ell[t] <- alpha * y[t] + (1 - alpha) * ell[t - 1]
}
tibble::tibble(t = 1:4, y = y, level = ell, forecast_next = ell)
```

You should get levels `10, 11, 11, 12`. Multi-step forecasts from $t=4$ are all `12`.

## Holt toy series

Data: $y = (10, 12, 14, 15)$, $\alpha = \beta^* = 0.5$, $\ell_0 = 10$, $b_0 = 2$.

```{r}
y <- c(10, 12, 14, 15)
alpha <- 0.5
beta <- 0.5
ell <- numeric(4)
b <- numeric(4)
ell_prev <- 10
b_prev <- 2
for (t in 1:4) {
  ell[t] <- alpha * y[t] + (1 - alpha) * (ell_prev + b_prev)
  b[t] <- beta * (ell[t] - ell_prev) + (1 - beta) * b_prev
  ell_prev <- ell[t]
  b_prev <- b[t]
}
tibble::tibble(t = 1:4, y = y, level = round(ell, 4), slope = round(b, 4),
               fcast_1 = round(ell + b, 4))
```

Check $t=1$: $\ell_1=11$, $b_1=1.5$. From $t=4$, the $h$-step forecast is $\ell_4 + h b_4$.

## Additive seasonal update (one step)

$y_5=22$, $\ell_4=18.75$, $s_1=1.25$, $\alpha=\gamma=0.5$:

```{r}
ell5 <- 0.5 * (22 - 1.25) + 0.5 * 18.75
s5 <- 0.5 * (22 - ell5) + 0.5 * 1.25
c(level = ell5, season_Q1 = s5, fcast_Q2 = ell5 - 8.75)
```

# Data: Australian beer production

```{r}
beer <- aus_production |>
  filter(!is.na(Beer)) |>
  select(Quarter, Beer)

beer |> autoplot(Beer) +
  labs(title = "Australian beer production", y = "megalitres")
```

**Interpret:** Trend? Seasonality? Changing variance? Additive or multiplicative season?

# Seasonal graphics

```{r}
beer |> gg_season(Beer) + labs(title = "Seasonal plot: Beer")
beer |> gg_subseries(Beer) + labs(title = "Subseries plot: Beer")
```

# Fit ETS (automatic)

```{r}
fit_auto <- beer |> model(ets = ETS(Beer))
report(fit_auto)
```

```{r}
fit_auto |>
  forecast(h = "3 years") |>
  autoplot(beer, level = NULL) +
  labs(title = "ETS automatic forecast")
```

# Compare specifications + AICc

```{r}
fit_cmp <- beer |>
  model(
    snaive = SNAIVE(Beer),
    ses = ETS(Beer ~ error("A") + trend("N") + season("N")),
    holt_winters = ETS(Beer ~ error("A") + trend("A") + season("A")),
    ets_auto = ETS(Beer)
  )

glance(fit_cmp) |> arrange(AICc)
accuracy(fit_cmp)
```

# Hold-out evaluation

```{r}
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)

fit_ho <- train |>
  model(
    snaive = SNAIVE(Beer),
    ses = ETS(Beer ~ error("A") + trend("N") + season("N")),
    auto = ETS(Beer)
  )

fc <- fit_ho |> forecast(h = nrow(test))
accuracy(fc, beer) |> arrange(RMSE)
fc |> autoplot(beer, level = NULL)
```

# Residuals

```{r}
fit_auto |> gg_tsresiduals()
```

# Stretch: employment SES vs Holt

```{r}
us_total <- us_employment |>
  filter(Title == "All Employees, Total Nonfarm", year(Month) >= 1990) |>
  select(Month, Employed)

us_total |>
  model(
    ses = ETS(Employed ~ error("A") + trend("N") + season("N")),
    holt = ETS(Employed ~ error("A") + trend("A") + season("N")),
    damped = ETS(Employed ~ error("A") + trend("Ad") + season("N"))
  ) |>
  forecast(h = 36) |>
  autoplot(us_total, level = NULL)
```

# Your turn

1. Refit Beer with multiplicative seasonality (if admissible) and compare AICc.
2. Hold out the last 8 quarters; compare `SNAIVE` vs `ETS` on RMSE / MAE.
3. Write two sentences recommending a model to a brewery planner.
