---
title: "Lab 01 — ETS and AIC"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(fpp3)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4
)
source("lab_submission.R")

beer <- aus_production |>
  filter(!is.na(Beer)) |>
  select(Quarter, Beer)

lab_meta <- list(id = "lab01", title = "ETS and AIC")
lab_prompts <- c(
  q1 = "SES: y=(10,12,11,13), α=0.5, ℓ₀=10. What is ℓ₄?",
  q2 = "Beer seasonal amplitude looks roughly stable. Prefer additive or multiplicative season?",
  q3 = "In the ETS taxonomy, SES is which triple (Error, Trend, Season)?",
  q4 = "Why should you not choose between ETS and ARIMA using AICc alone?",
  q5 = "In one or two sentences, recommend a Beer forecast method to a brewery planner (name the method and one reason)."
)
```

## What you will do

Reactivate pattern reading, check SES arithmetic, fit ETS with `fpp3`, and compare models with AICc **and** hold-out accuracy. At the end, download a `.txt` and upload it per LMS instructions.

Packages (Posit Cloud / local): `fpp3`, `learnr`. Open this file and click **Run Document**.

## SES warm-up

Data: $y=(10,12,11,13)$, $\alpha=0.5$, $\ell_0=10$. Complete / run the recursion.

```{r ses-ex, exercise=TRUE, exercise.lines=12}
y <- c(10, 12, 11, 13)
alpha <- 0.5
ell <- numeric(4)
ell0 <- 10
ell[1] <- alpha * y[1] + (1 - alpha) * ell0
for (t in 2:4) {
  ell[t] <- alpha * y[t] + (1 - alpha) * ell[t - 1]
}
tibble(t = 1:4, y = y, level = ell)
```

```{r q1}
question(
  lab_prompts[["q1"]],
  answer("12", correct = TRUE),
  answer("10"),
  answer("11"),
  answer("13"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

Multi-step SES forecasts from $t=4$ are **flat**: all equal to $\ell_4$.

## Beer patterns

```{r beer-plot, exercise=TRUE, exercise.lines=6}
beer |>
  autoplot(Beer) +
  labs(title = "Australian beer production", y = "megalitres")
```

```{r beer-season, exercise=TRUE, exercise.lines=5}
beer |> gg_season(Beer) + labs(title = "Seasonal plot: Beer")
```

```{r q2}
question(
  lab_prompts[["q2"]],
  answer("Additive (stable seasonal swings)", correct = TRUE),
  answer("Multiplicative (swings grow with level)"),
  answer("Neither — Beer has no seasonality"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

## Fit automatic ETS

```{r ets-fit, exercise=TRUE, exercise.lines=6}
fit_auto <- beer |> model(ets = ETS(Beer))
report(fit_auto)
```

```{r ets-fc, exercise=TRUE, exercise.lines=6}
fit_auto <- beer |> model(ets = ETS(Beer))
fit_auto |>
  forecast(h = "3 years") |>
  autoplot(beer, level = NULL)
```

```{r q3}
question(
  lab_prompts[["q3"]],
  answer("(A, N, N)", correct = TRUE),
  answer("(A, A, A)"),
  answer("(A, A, N)"),
  answer("(M, A, M)"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

## AICc vs hold-out

Compare candidates on the **same** Beer series. Sort by AICc, then check a hold-out.

```{r aicc-ex, exercise=TRUE, exercise.lines=14}
fit_cmp <- beer |>
  model(
    snaive = SNAIVE(Beer),
    ses = ETS(Beer ~ error("A") + trend("N") + season("N")),
    hw = ETS(Beer ~ error("A") + trend("A") + season("A")),
    auto = ETS(Beer)
  )
glance(fit_cmp) |> arrange(AICc)
```

```{r holdout-ex, exercise=TRUE, exercise.lines=14}
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)
fit_ho <- train |>
  model(
    snaive = SNAIVE(Beer),
    ses = ETS(Beer ~ error("A") + trend("N") + season("N")),
    auto = ETS(Beer)
  )
fc <- fit_ho |> forecast(h = nrow(test))
accuracy(fc, beer) |> arrange(RMSE)
```

```{r q4}
question(
  lab_prompts[["q4"]],
  answer("AICc is not a universal score across model families; use hold-out / TS CV to compare ETS vs ARIMA.", correct = TRUE),
  answer("AICc is only defined for seasonal naïve."),
  answer("Lower AICc always means better forecasts on every test set."),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

## Residuals (optional check)

```{r resid-ex, exercise=TRUE, exercise.lines=8}
beer |>
  model(ETS(Beer)) |>
  augment() |>
  filter(!is.na(.innov)) |>
  gg_tsdisplay(.innov, plot_type = "histogram")
```

Look for leftover seasonal spikes in the residual ACF.

## Recommendation

```{r q5}
question_text(
  lab_prompts[["q5"]],
  open_answer(),
  placeholder = "e.g. seasonal ETS / SNAIVE because … hold-out RMSE … residuals …",
  trim = TRUE
)
```

## Submit

Answer the questions above first, type your name, download the `.txt`, and upload it.

```{r submit-ui, echo=FALSE}
shiny::textInput("student_name", "Your name", placeholder = "Name Surname")
shiny::downloadButton("download_txt", "Download submission .txt")
```

```{r submit-server, context="server"}
output$download_txt <- shiny::downloadHandler(
  filename = function() lab_submission_filename(input$student_name, lab_meta$id),
  content = function(file) {
    write_lab_submission(
      path = file,
      name = input$student_name,
      lab_id = lab_meta$id,
      lab_title = lab_meta$title,
      state = learnr::get_tutorial_state(),
      prompts = lab_prompts
    )
  }
)
```
