---
title: "Session 6 — Practice Lab: ETS & ARIMA"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

> **Slides vs lab.** Theory and figures live on the slides. This lab is for runnable code.
>
> Use the same Beer hold-out as elsewhere: train through **2007 Q4**, test from **2008 Q1**.
> For Spain GDP exercises use train **Year <= 2010**.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

# Before you start

This lab has **5 exercises**. Solutions are **not** included here — work through
them in class (pairs/groups), then compare against the class debrief.
Each exercise follows the same pipeline used in Session 1--5:

> plot → transform? → difference? → ACF/PACF → fit ETS & ARIMA candidates → compare → diagnose

```{r}
library(fpp3)
```

---

# Exercise 1 — Multiplicative seasonality (a10 antidiabetic sales)

```{r}
a10 <- PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(Cost = sum(Cost)) |>
  mutate(Cost = Cost / 1e6)

a10 |> autoplot(Cost) +
  labs(title = "Australian antidiabetic drug sales", y = "$ million")
```

**Your turn:**

1. Look at the seasonal pattern. Does it grow with the level? What does that imply
   about additive vs. multiplicative seasonality (for ETS) or a log transform (for ARIMA)?
2. Fit `ETS(Cost)` (auto) and `ARIMA(log(Cost))` (auto). Report `report()` for both.
3. Use `glance()` to compare `AICc` and `BIC` **within** each model family (ETS vs. ETS,
   ARIMA vs. ARIMA — not across families).
4. Forecast 24 months ahead with each model and plot against history. Do the forecast
   shapes look plausible?
5. In 2–3 sentences: which model would you recommend for a 2-year pharmacy stocking
   forecast, and why?

```{r}
# Your code here

```

---

# Exercise 2 — Differencing and stationarity (Australian GDP)

```{r}
gdp <- global_economy |>
  filter(Code == "AUS") |>
  select(Year, GDP) |>
  mutate(log_gdp = log(GDP))

gdp |> autoplot(log_gdp) +
  labs(title = "Australia: log(GDP)")
```

**Your turn:**

1. Use `features(log_gdp, unitroot_ndiffs)` to check how many differences are needed.
2. Use `gg_tsdisplay(difference(log_gdp), plot_type = "partial")` to inspect the
   differenced series' ACF/PACF. Propose 1–2 candidate `(p, d, q)` combinations by eye.
3. Fit your candidates plus `ARIMA(log_gdp)` (auto-selected). Compare `AICc`.
4. Split off the last 5 years as a test set. Compare `RMSE` and `MAPE` **on the
   original GDP scale** (not log) for your best ARIMA model vs. `ETS(log_gdp)`.
5. Which model wins on AICc? Which wins on holdout accuracy? If they disagree, which
   would you trust more, and why?

```{r}
# Your code here

```

---

# Exercise 3 — Seasonal ARIMA vs. ETS with a changing pattern

```{r}
arrivals_jp <- aus_arrivals |>
  filter(Origin == "Japan")

arrivals_jp |> autoplot(Arrivals) +
  labs(title = "Arrivals to Australia from Japan")
```

**Your turn:**

1. Describe what changes over time in this series (trend direction, seasonal
   amplitude, any structural breaks you can see).
2. Fit `ETS(Arrivals)` and `ARIMA(Arrivals)` (auto, let it choose a transform/order).
3. Run `gg_tsresiduals()` for both. Does either show leftover autocorrelation?
4. Compare `accuracy()` on a holdout of the last 3 years.
5. Which model handles the structural change more gracefully? Justify using both the
   plot and the diagnostics.

```{r}
# Your code here

```

---

# Exercise 4 — Pure ARMA identification (Hudson Bay hare trappings)

```{r}
pelt |> autoplot(Hare) +
  labs(title = "Hudson Bay Company: hare pelt trappings", y = "Number trapped")
```

**Your turn:**

1. This series has no trend and no seasonality (it's annual). Is it already
   stationary? Check with `unitroot_kpss` if you like.
2. Plot `ACF(Hare)` and `PACF(Hare)`. Based on the shapes (cut off vs. decay),
   propose 1–2 candidate `(p, q)` pairs.
3. Fit your guesses with `ARIMA(Hare ~ pdq(p, 0, q))` for your chosen `p, q`, plus
   `ARIMA(Hare)` auto-selected. Compare `AICc`.
4. How close did your by-eye guess get to the auto-selected model?
5. The fitted AR order is related to the length of the population cycle. Roughly how
   many years does one hare population cycle last, based on the plot?

```{r}
# Your code here

```

---

# Exercise 5 (stretch) — Pick your own series

```{r}
aus_retail |> distinct(State, Industry) |> head(10)
```

**Your turn:**

1. Pick **one** `State` × `Industry` combination from `aus_retail` (filter to a single
   series with `filter(State == "...", Industry == "...")`).
2. Run the full pipeline: plot → transform? → difference? → ACF/PACF → fit `ETS()`
   and `ARIMA()` (both auto) → compare AICc within family → holdout accuracy → residuals.
3. Write your final recommendation in **3 sentences**, as if presenting to a retail
   category manager who has never heard of AICc.

```{r}
# Your code here

```

---

# Wrap-up

Be ready to share, for **one** of your exercises:

- the model you chose,
- the single plot you'd show a non-technical stakeholder,
- one thing that surprised you.
