---
title: "Lab 06 — Practice ETS and ARIMA"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(fpp3)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4
)
source("lab_submission.R")

lab_meta <- list(id = "lab06", title = "Practice ETS and ARIMA")
lab_prompts <- c(
  q1 = "AICc may be used to rank:",
  q2 = "Hudson Bay hare is annual. A seasonal ARIMA with m=4 is:",
  q3 = "If you model log(GDP), stakeholder RMSE should be computed on:",
  q4 = "Which practice series felt hardest, and what was the hard decision?"
)
```

## What you will do

Practice decisions before the midterm. Run one exercise in the session Rmd if you want the full bake-off; answer the decision questions here.

Open this file and click **Run Document**. Packages: `fpp3`, `learnr`.
Exercise chunks run when you click Run Code. They are not knitted in one pass.

```{r q1}
question(
  lab_prompts[["q1"]],
  answer("Models of the same series, transform, and family", correct = TRUE),
  answer("Any ETS against any ARIMA"),
  answer("A log model against a level model"),
  answer("The model with the most parameters"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
```{r q2}
question(
  lab_prompts[["q2"]],
  answer("Inappropriate — there is no quarterly season", correct = TRUE),
  answer("Required"),
  answer("The airline model"),
  answer("The same as SNAIVE"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
```{r q3}
question(
  lab_prompts[["q3"]],
  answer("The original GDP scale", correct = TRUE),
  answer("The log scale only"),
  answer("The AICc scale"),
  answer("First differences only"),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
```{r q4}
question_text(
  lab_prompts[["q4"]],
  open_answer(),
  placeholder = "Name the series and the choice you had to justify.",
  trim = TRUE
)
```

## Submit

Answer the questions above first, type your name, download the `.txt`, and upload it.

```{r submit-ui, echo=FALSE}
shiny::textInput("student_name", "Your name", placeholder = "Name Surname")
shiny::downloadButton("download_txt", "Download submission .txt")
```

```{r submit-server, context="server"}
output$download_txt <- shiny::downloadHandler(
  filename = function() lab_submission_filename(input$student_name, lab_meta$id),
  content = function(file) {
    write_lab_submission(
      path = file,
      name = input$student_name,
      lab_id = lab_meta$id,
      lab_title = lab_meta$title,
      state = learnr::get_tutorial_state(),
      prompts = lab_prompts
    )
  }
)
```
