---
title: "Session 8 — Recreating practice-mock R output"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

> **Slides vs lab.** Theory and figures live on the slides. This lab is for runnable code.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

# Purpose

This (optional) notebook recreates R output **in the same style** as
`readings/08-practice-midterm.md` Section C, so you can:

- see how `report()`, `glance()`/`accuracy()`, and residual diagnostics are actually
  produced (not just read as printed text), and
- practice generating and reading this kind of output yourself before the exam.

The exact numbers below will **not** match the practice mock exactly (different
series were used to build that document) — the point is to practice the *workflow*
of producing and reading this output, not to reverse-engineer the practice exam.

```{r}
library(fpp3)
```

---

# Recreating a `report()`-style output (like Section C1)

```{r}
fit_c1 <- aus_retail |>
  filter(State == "Victoria", Industry == "Department stores") |>
  summarise(Turnover = sum(Turnover)) |>
  model(ets = ETS(Turnover))

report(fit_c1)
```

**Practice:** without looking at the numbers above ahead of time, write down (i) the
ETS(E,T,S) form, (ii) what $\alpha$ tells you about level updating, (iii) which
information criterion you'd use to compare this to another ETS spec on the same
series. Then check yourself against the printed output.

---

# Recreating an `accuracy()`-style comparison (like Section C2)

```{r}
train_c2 <- aus_retail |>
  filter(State == "Victoria", Industry == "Department stores") |>
  summarise(Turnover = sum(Turnover)) |>
  filter(Month <= max(Month) - 24)

test_c2 <- aus_retail |>
  filter(State == "Victoria", Industry == "Department stores") |>
  summarise(Turnover = sum(Turnover))

fit_c2 <- train_c2 |>
  model(snaive = SNAIVE(Turnover), ets = ETS(Turnover), arima = ARIMA(Turnover))

fit_c2 |> forecast(h = 24) |> accuracy(test_c2)
```

**Practice:** which model wins on RMSE? On MAPE? What does the MASE column tell you
relative to the seasonal naive benchmark? (Same questions as Section C2 — different
series/numbers.)

---

# Recreating a residual-diagnostics check (like Section C3)

```{r}
fit_c2 |>
  select(arima) |>
  gg_tsresiduals()

augment(fit_c2) |>
  filter(.model == "arima") |>
  features(.innov, ljung_box, lag = 24, dof = 4)
```

**Practice:** based on the ACF plot and the Ljung–Box `p_value`, are these residuals
consistent with white noise? What would you do next if not?

---

# After you've tried it yourself

Compare your written answers against the instructor debrief. If your interpretation
of any of `report()`, `accuracy()`, or `gg_tsresiduals()` output didn't match, that's
exactly the kind of gap Section C of the exam is designed to probe — revisit
`readings/07-readings.md` before the mock.
