---
title: "Session 9 — Time series regression (1/2) (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

## Design matrix and fitted values

Two years of quarterly data, trend $t=1,\dots,8$, seasonal dummies $D_2,D_3,D_4$ (Q1 baseline).
Given coefficients (not estimated): $\beta_0=10,\ \beta_1=0.5,\ \beta_2=-2,\ \beta_3=-3,\ \beta_4=4$.

```{r}
quarters <- rep(1:4, times = 2)
toy <- tibble::tibble(
  t = 1:8,
  Quarter = quarters,
  D2 = as.numeric(quarters == 2),
  D3 = as.numeric(quarters == 3),
  D4 = as.numeric(quarters == 4)
)

beta <- c(Intercept = 10, t = 0.5, D2 = -2, D3 = -3, D4 = 4)

X <- as.matrix(cbind(Intercept = 1, toy[, c("t", "D2", "D3", "D4")]))
toy$fitted <- as.numeric(X %*% beta)
toy
```

You should get fitted values `10.5, 9, 8.5, 16, 12.5, 11, 10.5, 18` — matching the slide table exactly.
Check the trend-only step: $\hat y_8-\hat y_4 = 4\beta_1 = 2$.

# Data: Australian beer production

```{r}
beer <- aus_production |>
  filter(!is.na(Beer)) |>
  select(Quarter, Beer)

beer |> autoplot(Beer) +
  labs(title = "Australian beer production", y = "megalitres")
```

# Fit TSLM (trend + season) on the full series

```{r}
fit <- beer |>
  model(
    snaive = SNAIVE(Beer),
    tslm = TSLM(Beer ~ trend() + season()),
    ets = ETS(Beer)
  )

report(fit |> select(tslm))
glance(fit) |> arrange(AICc)
```

```{r}
fit |> select(tslm) |> gg_tsresiduals()
```

```{r}
fit |> forecast(h = "2 years") |> autoplot(beer, level = NULL) +
  labs(title = "TSLM vs ETS vs SNAIVE")
```

# Hold-out evaluation: train through 2007 Q4, test from 2008 Q1

```{r}
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)

fit_ho <- train |>
  model(
    snaive = SNAIVE(Beer),
    tslm = TSLM(Beer ~ trend() + season()),
    ets = ETS(Beer)
  )

fc_ho <- fit_ho |> forecast(h = nrow(test))
accuracy(fc_ho, beer) |> arrange(RMSE)
fc_ho |> autoplot(beer, level = NULL) +
  labs(title = "Hold-out forecasts (train through 2007 Q4)")
```

# Your turn

1. Add only `trend()` (no season) and compare residual seasonality with `gg_tsresiduals()`.
2. Interpret the largest seasonal coefficient (`report()`) in one sentence, relative to the omitted quarter.
3. From the hold-out table above, which model wins on RMSE? Does the winner match the in-sample AICc ranking?
