---
title: "Session 12 — Dynamic regression (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

## Toy distributed-lag model

$y_t = 0.5 + 0.3 x_t + 0.2 x_{t-1}$, $x_0 = 1$, $x = (1, 2, 1.5, 2.5)$.

```{r}
x <- c(1, 2, 1.5, 2.5)
x_lag <- dplyr::lag(x, default = 1)  # x0 = 1 (pre-sample)
y <- 0.5 + 0.3 * x + 0.2 * x_lag
tibble::tibble(t = 1:4, x = x, x_lag = x_lag, y = y)
```

Expected `y`: `1.00, 1.30, 1.35, 1.55`. Total eventual effect of a permanent $+1$ change in
Income: $0.3 + 0.2 = 0.5$.

# Data: US consumption & income

```{r}
us_change |> autoplot(vars(Consumption, Income)) +
  labs(title = "US consumption & income growth")
```

# TSLM vs regression with ARIMA errors

```{r}
fit <- us_change |>
  model(
    lm = TSLM(Consumption ~ Income),
    dyn = ARIMA(Consumption ~ Income + PDQ(0, 0, 0))
  )
glance(fit) |> select(.model, AICc)
report(fit |> select(dyn))
# Coefficients as numbers (match the slide table --- no barplot)
tidy(fit |> select(dyn)) |>
  filter(term != "sigma2") |>
  select(term, estimate, std.error, statistic, p.value)
```

```{r}
fit |> select(lm) |> gg_tsresiduals()
fit |> select(dyn) |> gg_tsresiduals()
```

Compare the two residual ACF panels: `dyn` should look much closer to white noise than `lm`.

# Forecasting with a future-Income path

We assume Income holds at its recent average for the next 8 quarters (a simple,
explicit scenario - not a prediction) and forecast Consumption from the dynamic
regression model.

```{r}
future_income <- new_data(us_change, 8) |>
  mutate(Income = mean(tail(us_change$Income, 8)))

fc_dyn <- forecast(fit |> select(dyn), new_data = future_income)
fc_dyn

fc_dyn |>
  autoplot(us_change |> filter_index("2015 Q1" ~ .), level = NULL) +
  labs(title = "Dynamic regression forecast (Income at recent average)")
```

# Your turn

1. Add `Savings` to the dynamic regression (`ARIMA(Consumption ~ Income + Savings + PDQ(0,0,0))`);
   does the residual ACF improve further?
2. Build a second future-Income scenario (e.g. $+1\%$ per quarter compounding) and compare the
   resulting forecast band with the "hold flat" scenario above.
3. Outline Assignment A2 for your group: which regressors, what lag structure, what future-$x$
   assumption will you use?
