---
title: "Session 3 — ARMA processes (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.  
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

## AR(1) recursion

$\phi = 0.5$, $y_0 = 0$, $\varepsilon = (2, -1, 1.5, -0.5)$.

```{r}
phi <- 0.5
eps <- c(2, -1, 1.5, -0.5)
y <- numeric(4)
y[1] <- phi * 0 + eps[1]
for (t in 2:4) y[t] <- phi * y[t - 1] + eps[t]
tibble::tibble(t = 1:4, eps = eps, y = y)
```

Expected $y$: `2, 0, 1.5, 0.25`.

## MA(1) ACF formula

```{r}
theta <- 0.8
rho1 <- theta / (1 + theta^2)
c(theta = theta, rho1 = rho1)
```

Expected $\rho_1 \approx 0.488$.

## MA(1) path

$\theta = 0.5$, $\varepsilon = (1, -2, 0.5, 1)$.

```{r}
theta <- 0.5
eps <- c(1, -2, 0.5, 1)
y <- eps + theta * dplyr::lag(eps, default = 0)
tibble::tibble(t = 1:4, eps = eps, y = y)
```

Expected $y$: `1, -1.5, -0.5, 1.25`.

# Simulate AR(1) and MA(1)

```{r}
set.seed(42)
n <- 300
e <- rnorm(n)

phi <- 0.8
yar <- numeric(n)
for (t in 2:n) yar[t] <- phi * yar[t - 1] + e[t]

theta <- 0.8
yma <- e + theta * dplyr::lag(e, default = 0)

sim <- tibble(index = 1:n, AR1 = yar, MA1 = yma) |>
  as_tsibble(index = index)

sim |> gg_tsdisplay(AR1, plot_type = "partial") + labs(title = "AR(1), phi = 0.8")
sim |> gg_tsdisplay(MA1, plot_type = "partial") + labs(title = "MA(1), theta = 0.8")
```

# Persistence: phi = 0.4 vs 0.9 (same shocks)

```{r}
set.seed(42)
e <- rnorm(200)
sim_phi <- function(phi) {
  y <- numeric(200)
  for (t in 2:200) y[t] <- phi * y[t - 1] + e[t]
  y
}
tibble(index = 1:200, lo = sim_phi(0.4), hi = sim_phi(0.9)) |>
  as_tsibble(index = index) |>
  pivot_longer(-index) |>
  autoplot(value) +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(title = "Same shocks; different phi", y = NULL)
```

# Fit ARMA to Canadian lynx

```{r}
pelt |> autoplot(Lynx) + labs(title = "Canadian lynx (pelt)")
fit <- pelt |> model(
  ar2 = ARIMA(Lynx ~ pdq(2, 0, 0) + PDQ(0, 0, 0)),
  arma11 = ARIMA(Lynx ~ pdq(1, 0, 1) + PDQ(0, 0, 0)),
  auto = ARIMA(Lynx ~ PDQ(0, 0, 0))
)
glance(fit) |> arrange(AICc)
fit |> select(auto) |> gg_tsresiduals()
```

# Your turn

## AR(2) simulation and PACF cutoff

```{r}
set.seed(7)
n <- 300
e <- rnorm(n)
phi1 <- 1.3
phi2 <- -0.7
y2 <- numeric(n)
for (t in 3:n) {
  y2[t] <- phi1 * y2[t - 1] + phi2 * y2[t - 2] + e[t]
}
tibble(index = 1:n, AR2 = y2) |>
  as_tsibble(index = index) |>
  gg_tsdisplay(AR2, plot_type = "partial") +
  labs(title = "AR(2): expect PACF cutoff after lag 2")
```

1. Blind identification: partner simulates AR or MA; you guess from ACF/PACF.
2. For `pelt$Lynx`, confirm AICc ranking of AR(2) vs ARMA(1,1) vs auto (above).
