---
title: "Session 2 — Stationarity and differencing"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

> **Slides vs lab.** Theory and figures live on the slides. This lab is for runnable code.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

```{r}
library(fpp3)
```


# Nonstationary example: Alphabet (GOOG) closing price

Daily USD closing share price of Alphabet Inc.\ (ticker `GOOG`),
from `gafa_stock` in `{tsibbledata}`.

```{r}
goog <- gafa_stock |>
  filter(Symbol == "GOOG") |>
  filter(year(Date) >= 2016) |>
  select(Date, Close)

goog |> autoplot(Close) + labs(title = "Alphabet (GOOG) closing price", y = "USD")
goog |> gg_tsdisplay(Close, plot_type = "partial")
```

# Difference once

```{r}
goog |>
  mutate(diff_close = difference(Close)) |>
  gg_tsdisplay(diff_close, plot_type = "partial")
```

# Seasonal example: beer

```{r}
beer <- aus_production |> filter(!is.na(Beer))
beer |> features(Beer, list(ndiffs = unitroot_ndiffs, nsdiffs = unitroot_nsdiffs))

beer_d <- beer |>
  mutate(Beer_d = difference(Beer, lag = 4)) |>
  filter(!is.na(Beer_d))

beer_d |> gg_tsdisplay(Beer_d, plot_type = "partial")
```

# Hand calculations (match the slides)

Using Beer 1992–1993 from the slides:

```{r}
beer_toy <- aus_production |>
  filter(Quarter >= yearquarter("1992 Q1"),
         Quarter <= yearquarter("1993 Q4")) |>
  select(Quarter, Beer)

beer_toy
```

1. Compute $\nabla y_t$, $\nabla_4 y_t$, and $(1-B)(1-B^4)y_t$ by hand for 1993.
2. Check with:

```{r}
beer_toy |>
  mutate(
    d1 = difference(Beer),
    d4 = difference(Beer, lag = 4),
    d14 = difference(difference(Beer, lag = 4))
  )
```

# Backshift practice

Expand $(1-B)^2 y_t$ and $(1-B)(1-\tfrac12 B)y_t$; evaluate on a short toy vector and verify in R.

# Your turn

1. For `us_employment` (one industry of your choice), decide whether to log, difference, and/or seasonally difference. Justify with plots + `unitroot_*`.
2. Over-difference once more than needed and show how the ACF changes.
