---
title: "Session 7 — Timed Practice Problems"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

> **Slides vs lab.** Theory and figures live on the slides. This lab is for runnable code.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

# How to use this lab

Simulate exam conditions: set a timer for each problem, work without looking up
documentation, and write your answer as a comment **before** checking it runs.
Solutions are not included here - see the instructor debrief / posted notes after
Session 7.

Fixed splits used elsewhere in the course (for consistency):
Beer train through `2007 Q4`; Spain GDP train `Year <= 2010`.

```{r}
library(fpp3)
```

---

# Problem 1 (10 min) — Reading a report() output

```{r}
fit <- aus_production |>
  filter(!is.na(Cement)) |>
  model(ets = ETS(Cement))

report(fit)
```

**Answer without re-running the code differently:**

1. What is the fitted ETS(E,T,S) form, in words (error/trend/season)?
2. Name the two smoothing parameters shown and, in one sentence each, what a value
   close to 1 vs. close to 0 would mean for that parameter.
3. Is `AICc` or `sigma2` the quantity you'd use to compare this model against a
   *different* ETS specification on the same series?

```{r}
# Your written answers as comments

```

---

# Problem 2 (10 min) — Differencing decision

```{r}
h02 <- global_economy |>
  filter(Code == "USA") |>
  select(Year, GDP) |>
  mutate(log_gdp = log(GDP))

h02 |>
  features(log_gdp, unitroot_ndiffs)

h02 |> gg_tsdisplay(difference(log_gdp), plot_type = "partial")
```

**Answer:**

1. How many ordinary differences does `unitroot_ndiffs` recommend?
2. Based on the ACF/PACF of the differenced series, propose one candidate
   `(p, d, q)`.
3. Write the fitted model equation in backshift-operator notation for your proposal
   (use $c$ for a constant/drift term if relevant).

```{r}
# Your written answers as comments

```

---

# Problem 3 (10 min) — AICc table under time pressure

```{r}
fit3 <- aus_production |>
  filter(!is.na(Beer)) |>
  model(
    snaive    = SNAIVE(Beer),
    ets_auto  = ETS(Beer),
    arima_auto = ARIMA(Beer)
  )

glance(fit3) |> select(.model, AICc, BIC)
```

**Answer:**

1. Why can't you directly compare the ETS row's AICc to the ARIMA row's AICc here?
2. What *can* you compare across all three rows to make a fair final decision?
3. Compute that fair comparison (write the code) and report the winner.

```{r}
# Your code + written answer

```

---

# Problem 4 (10 min) — ARMA identification speed round

```{r}
pelt |> gg_tsdisplay(Lynx, plot_type = "partial")
```

**Answer:**

1. From the ACF/PACF shapes alone, propose an AR order, MA order, or say "ARMA,
   can't tell by eye — let `ARIMA()` search."
2. Fit your proposal and `ARIMA(Lynx)` (auto). How far off was your guess on AICc?

```{r}
# Your code + written answer

```

---

# Problem 5 (10 min) — Interpret accuracy() output

```{r}
train5 <- aus_production |> filter(!is.na(Beer), Quarter <= max(Quarter) - 8)
test5  <- aus_production |> filter(!is.na(Beer))

fit5 <- train5 |> model(ets = ETS(Beer), arima = ARIMA(Beer), snaive = SNAIVE(Beer))

fit5 |> forecast(h = 8) |> accuracy(test5)
```

**Answer:**

1. Which model has the lowest test-set RMSE? The lowest MAPE? Are they the same model?
2. If RMSE and MAPE disagree on the "best" model, which would you report to a
   stakeholder who cares about percentage error, and why?
3. Is beating `SNAIVE` on this holdout enough, by itself, to justify using the more
   complex model in production? What else would you want to check?

```{r}
# Your written answers as comments

```

---

# Self-check

Total suggested time: ~50 minutes for 5 problems (matches roughly two-thirds of the
mock's calculation/R-output time budget). Note which problems took longer than
their suggested time — that's exactly where to focus remaining study time.
