---
title: "Session 10 — Time series regression (2/2) (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

## Nested models: from AICc to relative likelihood

```{r}
fit <- us_change |>
  model(
    lm1 = TSLM(Consumption ~ Income),
    lm2 = TSLM(Consumption ~ Income + Savings + Unemployment),
    lm3 = TSLM(Consumption ~ Income + Savings)
  )
aicc_tab <- glance(fit) |> select(.model, AICc) |> arrange(AICc)
aicc_tab
```

```{r}
best <- min(aicc_tab$AICc)
aicc_tab |>
  mutate(delta_AICc = AICc - best,
         rel_likelihood = exp(-0.5 * delta_AICc))
```

You should see `lm2` as best (`delta_AICc = 0`, `rel_likelihood = 1`), `lm3` with
`delta_AICc ~ 18.5` (`rel_likelihood ~ 9.4e-05`), and `lm1` essentially ruled out.

# Data: US consumption growth

```{r}
us_change |> autoplot(Consumption) + labs(title = "US consumption growth")
```

```{r}
report(fit |> select(lm2))
fit |> select(lm2) |> gg_tsresiduals()
```

# Scenario forecasts: Income +/-1% for the next 8 quarters

We hold Income at $\pm 1\%$ around its recent average and forecast Consumption from the
best nested model (`lm2` uses Savings and Unemployment too; here we illustrate the
Income-only scenario mechanics with a simpler model so only Income needs an assumption).

```{r}
fit_income <- us_change |> model(lm = TSLM(Consumption ~ Income))

base_income <- mean(tail(us_change$Income, 8))

future_scenarios <- scenarios(
  Increase = new_data(us_change, 8) |> mutate(Income = 1.01 * base_income),
  Decrease = new_data(us_change, 8) |> mutate(Income = 0.99 * base_income),
  names_to = "Scenario"
)

fc_scenarios <- forecast(fit_income, new_data = future_scenarios)
fc_scenarios

fc_scenarios |>
  autoplot(level = NULL) +
  autolayer(us_change |> filter_index("2015 Q1" ~ .), Consumption) +
  labs(title = "Consumption forecasts under \u00b11% Income scenarios",
       y = "Consumption growth (%)")
```

# Spurious regression demonstration (`set.seed(1)`)

Two **independent** random walks; regress one on the other and inspect the (misleading)
significance plus the residual ACF tell-tale.

```{r}
set.seed(1)
n <- 100
x <- cumsum(rnorm(n))
y <- cumsum(rnorm(n))

spurious_fit <- lm(y ~ x)
summary(spurious_fit)
```

```{r}
resid_acf1 <- acf(residuals(spurious_fit), plot = FALSE)$acf[2]
c(R2 = summary(spurious_fit)$r.squared, resid_acf1 = resid_acf1)
```

```{r}
tibble(t = 1:n, x = x, y = y) |>
  as_tsibble(index = t) |>
  pivot_longer(c(x, y)) |>
  autoplot(value) +
  labs(title = "Two independent random walks (seed = 1)", y = NULL, x = "t")
```

Expect $R^2 \approx 0.21$, slope $t$-value $\approx -5.1$ ($p \ll 0.001$), and residual
lag-1 ACF $\approx 0.96$ — despite `x` and `y` being generated with no relationship at all.

# Your turn

1. Drop `Unemployment` from `lm2`; confirm the AICc/relative-likelihood ranking changes as expected.
2. Re-run the spurious-regression demo with `set.seed(2)`. Does the sign of the slope change? Does the
   qualitative message (high residual ACF) still hold?
3. Write two sentences to a stakeholder explaining why the $\pm1\%$ Income scenarios are assumptions,
   not predictions.
