---
title: "Session 0 — Time Series Analysis review (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Slides review TSA concepts with figures.  
> This lab rebuilds the prior-course workflow in `fpp3`.

# Packages

```{r}
library(fpp3)
```

# 1. Graphics checklist

```{r}
beer <- aus_production |>
  filter(!is.na(Beer)) |>
  select(Quarter, Beer)

beer |> autoplot(Beer)
beer |> gg_season(Beer)
beer |> gg_subseries(Beer)
beer |> ACF(Beer) |> autoplot()
```

**Write 3 bullets:** trend? season? ACF message?

# 2. STL decomposition

```{r}
beer |>
  model(stl = STL(Beer)) |>
  components() |>
  autoplot()
```

# 3. Benchmarks + accuracy

```{r}
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)

fit <- train |>
  model(
    mean = MEAN(Beer),
    naive = NAIVE(Beer),
    snaive = SNAIVE(Beer),
    drift = RW(Beer ~ drift())
  )

fc <- fit |> forecast(h = nrow(test))
accuracy(fc, beer) |> arrange(RMSE)
fc |> autoplot(beer, level = NULL)
fit |> select(snaive) |> gg_tsresiduals()
```

# 4. SES mini hand-check (bridge to Session 1)

```{r}
y <- c(10, 12, 11)
alpha <- 0.5
ell0 <- 10
ell <- numeric(3)
ell[1] <- alpha * y[1] + (1 - alpha) * ell0
for (t in 2:3) ell[t] <- alpha * y[t] + (1 - alpha) * ell[t - 1]
ell  # expect 10, 11, 11
```

# 5. Optional: multiplicative season candidate

```{r}
aus_production |>
  filter(!is.na(Gas)) |>
  autoplot(Gas) +
  labs(title = "Gas production")
```

# Your turn

1. Pick another series in `tsibbledata` and complete the graphics checklist.
2. Which benchmark would you defend for that series, and why?
3. List two TSA gaps you want Session 1–2 to reinforce.
