---
title: "Lab 0 — TSA review (interactive)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    language: en
runtime: shiny_prerendered
description: "Interactive fpp3 review: graphics, STL, benchmarks, SES hand-check."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(fpp3)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.2
)
gradethis::gradethis_setup()

beer <- aus_production |>
  filter(!is.na(Beer)) |>
  select(Quarter, Beer)
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)
```

## Welcome

> Same workflow as `00-tsa-review.Rmd`, but with **code exercises** and short quizzes.
> Run locally with:
>
> ```r
> learnr::run_tutorial(
>   "00-tsa-review-learnr",
>   # or open this .Rmd in RStudio and click Run Document
> )
> ```

You need package **`fpp3`** (pulls in tsibble / fable / feasts).

At the end (**Done**), download the `.txt` receipt and **submit it to Blackboard**.

```{r check-pkgs}
quiz(
  question(
    "Have you already installed and loaded `fpp3` successfully on your machine?",
    answer("Yes", correct = TRUE),
    answer("Not yet — I'll run `install.packages(\"fpp3\")` first"),
    allow_retry = TRUE
  )
)
```

## 1. Graphics checklist

Build the beer series and the four standard plots.

```{r ex-graphics, exercise = TRUE, exercise.lines = 12}
# beer is already in the environment
# 1) autoplot  2) gg_season  3) gg_subseries  4) ACF + autoplot

```

```{r ex-graphics-solution}
beer |> autoplot(Beer)
beer |> gg_season(Beer)
beer |> gg_subseries(Beer)
beer |> ACF(Beer) |> autoplot()
```

```{r ex-graphics-check}
grade_this({
  pass_if_equal(
    .result,
    beer |> ACF(Beer) |> autoplot(),
    message = "Nice — ACF plot is a good last step. Also run autoplot / season / subseries if you skipped them."
  )
  # Accept any plot / invisible result if code ran without error
  pass("Graphics checklist complete (or close enough). Look for trend, season, and ACF decay/spikes.")
})
```

```{r q-acf}
quiz(
  question(
    "For a strong quarterly seasonal series like beer, the ACF typically shows:",
    answer("Near-zero correlations at all nonzero lags"),
    answer("Spikes near seasonal lags (e.g. 4, 8, 12) and/or slow seasonal pattern", correct = TRUE),
    answer("Only lag-1 correlation and nothing else"),
    answer("Exactly 1 at every lag"),
    allow_retry = TRUE
  )
)
```

**Write (for yourself):** 3 bullets — trend? season? ACF message?

## 2. STL decomposition

```{r ex-stl, exercise = TRUE, exercise.lines = 8}
# Fit STL on Beer and autoplot the components

```

```{r ex-stl-solution}
beer |>
  model(stl = STL(Beer)) |>
  components() |>
  autoplot()
```

```{r ex-stl-check}
grade_this({
  pass("STL components plotted. Check which piece dominates: trend vs season vs remainder.")
})
```

## 3. Benchmarks + accuracy

Hold out from **2008 Q1** onward. Fit mean / naïve / seasonal naïve / drift; compare RMSE.

```{r ex-bench, exercise = TRUE, exercise.lines = 18}
# train and test are available
# model MEAN, NAIVE, SNAIVE, RW(~drift()), forecast, accuracy, arrange by RMSE

```

```{r ex-bench-solution}
fit <- train |>
  model(
    mean = MEAN(Beer),
    naive = NAIVE(Beer),
    snaive = SNAIVE(Beer),
    drift = RW(Beer ~ drift())
  )
fc <- fit |> forecast(h = nrow(test))
accuracy(fc, beer) |> arrange(RMSE)
```

```{r ex-bench-check}
grade_this({
  pass_if(~ is.data.frame(.result) || inherits(.result, "tbl_df"),
          message = "Good — inspect which method has the lowest RMSE.")
  pass("Run the accuracy table and note the winner. Then plot forecasts and snaive residuals.")
})
```

Optional plots (solution code):

```{r bench-plots, exercise = TRUE, exercise.lines = 10}
fit <- train |>
  model(
    mean = MEAN(Beer),
    naive = NAIVE(Beer),
    snaive = SNAIVE(Beer),
    drift = RW(Beer ~ drift())
  )
fc <- fit |> forecast(h = nrow(test))
# autoplot forecasts; gg_tsresiduals for snaive

```

```{r bench-plots-solution}
fit <- train |>
  model(
    mean = MEAN(Beer),
    naive = NAIVE(Beer),
    snaive = SNAIVE(Beer),
    drift = RW(Beer ~ drift())
  )
fc <- fit |> forecast(h = nrow(test))
fc |> autoplot(beer, level = NULL)
fit |> select(snaive) |> gg_tsresiduals()
```

```{r q-snaive}
quiz(
  question(
    "For quarterly beer with clear seasonality, which benchmark is usually the strongest starting point?",
    answer("MEAN (global average)"),
    answer("NAIVE (last value)"),
    answer("SNAIVE (same season last year)", correct = TRUE),
    answer("A random shuffle of the training set"),
    allow_retry = TRUE
  )
)
```

## 4. SES hand-check (bridge to Session 1)

SES level update: $\ell_t = \alpha y_t + (1-\alpha)\ell_{t-1}$.

Given $y = (10, 12, 11)$, $\alpha = 0.5$, $\ell_0 = 10$, compute $\ell_1, \ell_2, \ell_3$.

```{r ex-ses, exercise = TRUE, exercise.lines = 12}
y <- c(10, 12, 11)
alpha <- 0.5
ell0 <- 10
# fill ell; expect c(10, 11, 11)

```

```{r ex-ses-solution}
y <- c(10, 12, 11)
alpha <- 0.5
ell0 <- 10
ell <- numeric(3)
ell[1] <- alpha * y[1] + (1 - alpha) * ell0
for (t in 2:3) ell[t] <- alpha * y[t] + (1 - alpha) * ell[t - 1]
ell
```

```{r ex-ses-check}
grade_this({
  pass_if_equal(c(10, 11, 11), message = "Correct: levels are 10, 11, 11.")
  fail_if(~ !is.numeric(.result) || length(.result) != 3,
          message = "Return a length-3 numeric vector `ell`.")
  fail("Check the recursion: start from ell0 = 10 with alpha = 0.5.")
})
```

```{r q-ses}
quiz(
  question(
    "SES point forecasts from the last level $\\ell_T$ are:",
    answer("Flat: $\\hat{y}_{T+h|T} = \\ell_T$ for all $h \\ge 1$", correct = TRUE),
    answer("Always linearly trending"),
    answer("Always equal to the sample mean"),
    answer("Seasonally repeating the last year only"),
    allow_retry = TRUE
  )
)
```

## 5. Optional — multiplicative season?

```{r gas-plot}
aus_production |>
  filter(!is.na(Gas)) |>
  autoplot(Gas) +
  labs(title = "Gas production")
```

```{r q-gas}
quiz(
  question(
    "Looking at Gas, seasonal swings that grow with the level suggest:",
    answer("Additive seasonality only"),
    answer("Multiplicative seasonality (or a variance-stabilizing transform) is more plausible", correct = TRUE),
    answer("No seasonality at all"),
    answer("That MAE equals RMSE by definition"),
    allow_retry = TRUE
  )
)
```

## Your turn

1. Pick another series in `tsibbledata` and repeat the graphics checklist (new exercise box below).
2. Which benchmark would you defend for that series, and why?
3. List two TSA gaps you want Session 1–2 to reinforce.

```{r ex-your-turn, exercise = TRUE, exercise.lines = 14}
# Example idea: aus_production Gas, or another tsibble series
# autoplot / gg_season / gg_subseries / ACF

```

```{r ex-your-turn-solution}
aus_production |>
  filter(!is.na(Gas)) |>
  ACF(Gas) |>
  autoplot()
```

```{r ex-your-turn-check}
grade_this({
  pass("Good exploration. Bring your 3 bullets + preferred benchmark to class.")
})
```

## Done

1. Enter your details below.
2. Click **Download completion .txt for Blackboard**.
3. **Upload that `.txt` to Blackboard** (Lab 0 assignment / dropbox).

Inside this tutorial, quizzes/exercises are already checked by **gradethis**. Upload the `.txt` receipt to Blackboard as instructed.

```{r receipt-ui, context="render"}
shiny::tagList(
  shiny::textInput("stu_name", "Full name", placeholder = "Given name + surname"),
  shiny::textInput("stu_id", "Student ID or IE email", placeholder = "xxx@student.ie.edu"),
  shiny::textInput("stu_section", "Section / group (optional)", placeholder = "PPLEDBA"),
  shiny::br(),
  shiny::downloadButton("dl_lab00", "Download completion .txt for Blackboard")
)
```

```{r receipt-server, context="server"}
`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0 || !nzchar(as.character(x)[1])) y else x
}

graded_quizzes <- c("q-acf", "q-snaive", "q-ses", "q-gas")
exercise_labs <- c(
  "ex-graphics", "ex-stl", "ex-bench", "ex-ses", "ex-your-turn", "bench-plots"
)

output$dl_lab00 <- shiny::downloadHandler(
  filename = function() {
    nm <- gsub("[^A-Za-z0-9._-]+", "_", trimws(input$stu_name %||% "student"))
    if (!nzchar(nm)) nm <- "student"
    paste0("Lab00_TSA_Review_", nm, ".txt")
  },
  content = function(file) {
    name <- trimws(input$stu_name %||% "")
    shiny::validate(shiny::need(nzchar(name), "Enter your full name before downloading."))

    state <- shiny::isolate(
      shiny::reactiveValuesToList(learnr::get_tutorial_state())
    )
    info <- tryCatch(
      learnr::get_tutorial_info(),
      error = function(e) list(tutorial_id = "00-tsa-review-learnr")
    )

    lines_progress <- character()
    quiz_hit <- 0L
    quiz_ok <- 0L
    ex_hit <- 0L
    if (length(state)) {
      for (lab in sort(names(state))) {
        item <- state[[lab]]
        ans <- paste(item$answer %||% "", collapse = " | ")
        if (nchar(ans) > 80) ans <- paste0(substr(ans, 1, 77), "...")
        corr <- item$correct
        lines_progress <- c(
          lines_progress,
          sprintf(
            "- %s [%s] correct=%s @ %s",
            lab,
            item$type %||% "?",
            as.character(corr %||% NA),
            item$timestamp %||% ""
          ),
          if (nzchar(ans)) sprintf("    answer: %s", ans) else NULL
        )
        if (lab %in% graded_quizzes) {
          quiz_hit <- quiz_hit + 1L
          if (isTRUE(corr)) quiz_ok <- quiz_ok + 1L
        }
        if (lab %in% exercise_labs) ex_hit <- ex_hit + 1L
      }
    } else {
      lines_progress <- "- (no exercise/quiz submissions recorded in this session)"
    }

    n_quiz <- length(graded_quizzes)
    stamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
    txt <- c(
      "Lab 0 — TSA review (learnr) — completion receipt",
      "Course: Forecasting for Time Series (PPLEDBA), Fall 2026",
      sprintf("Tutorial: %s", info$tutorial_id %||% "00-tsa-review-learnr"),
      "SUBMIT_TO: Blackboard (Lab 0 assignment)",
      "",
      sprintf("Full name: %s", name),
      sprintf("Student ID / email: %s", trimws(input$stu_id %||% "")),
      sprintf("Section / group: %s", trimws(input$stu_section %||% "")),
      sprintf("Exported at: %s", stamp),
      sprintf("R version: %s", paste(R.version$major, R.version$minor, sep = ".")),
      "",
      sprintf(
        "SCORE_QUIZ: %d/%d (attempted %d of %d: %s)",
        quiz_ok, n_quiz, quiz_hit, n_quiz, paste(graded_quizzes, collapse = ", ")
      ),
      sprintf("SCORE_EXERCISES_SUBMITTED: %d/%d", ex_hit, length(exercise_labs)),
      "",
      "Progress (submitted items):",
      lines_progress,
      "",
      "I completed Lab 0 (interactive). I will upload this .txt to Blackboard."
    )
    writeLines(txt, file, useBytes = TRUE)
  }
)
```

Companion static lab: [`00-tsa-review.Rmd`](00-tsa-review.Rmd).
