---
title: "Lab 0 — TSA review (interactive)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    language: en
runtime: shiny_prerendered
description: "Interactive fpp3 review: graphics, STL, benchmarks, SES hand-check."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(fpp3)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.2
)
gradethis::gradethis_setup()

beer <- aus_production |>
  filter(!is.na(Beer)) |>
  select(Quarter, Beer)
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)
```

## Welcome

> Same workflow as `00-tsa-review.Rmd`, but with **code exercises** and short quizzes.

**How to run (do not open the `.html` file):** in RStudio open this `.Rmd` and click **Run Document**, or:

```r
install.packages(c("learnr", "remotes", "fpp3"))
# install.packages("remotes")
remotes::install_github("rstudio/gradethis")
rmarkdown::run("00-tsa-review-learnr.Rmd")
```

You need **`fpp3`**, **`learnr`**, and **`gradethis`** (from GitHub, as above).

At the end (**Done**), export a **`.txt`** receipt and submit it to Blackboard — **not** the `.html` file.

```{r check-pkgs}
quiz(
  question(
    "Have you already installed and loaded `fpp3` successfully on your machine?",
    answer("Yes", correct = TRUE),
    answer("Not yet — I'll run `install.packages(\"fpp3\")` first"),
    allow_retry = TRUE
  )
)
```

## 1. Graphics checklist

Build the beer series and the four standard plots.

```{r ex-graphics, exercise = TRUE, exercise.lines = 12}
# beer is already in the environment
# 1) autoplot  2) gg_season  3) gg_subseries  4) ACF + autoplot

```

```{r ex-graphics-solution}
beer |> autoplot(Beer)
beer |> gg_season(Beer)
beer |> gg_subseries(Beer)
beer |> ACF(Beer) |> autoplot()
```

```{r ex-graphics-check}
grade_this({
  # Any successful run counts; students should produce the four plots.
  pass("Graphics checklist complete (or close enough). Look for trend, season, and ACF decay/spikes.")
})
```

```{r q-acf}
quiz(
  question(
    "For a strong quarterly seasonal series like beer, the ACF typically shows:",
    answer("Near-zero correlations at all nonzero lags"),
    answer("Spikes near seasonal lags (e.g. 4, 8, 12) and/or slow seasonal pattern", correct = TRUE),
    answer("Only lag-1 correlation and nothing else"),
    answer("Exactly 1 at every lag"),
    allow_retry = TRUE
  )
)
```

**Write (for yourself):** 3 bullets — trend? season? ACF message?

## 2. STL decomposition

```{r ex-stl, exercise = TRUE, exercise.lines = 8}
# Fit STL on Beer and autoplot the components

```

```{r ex-stl-solution}
beer |>
  model(stl = STL(Beer)) |>
  components() |>
  autoplot()
```

```{r ex-stl-check}
grade_this({
  pass("STL components plotted. Check which piece dominates: trend vs season vs remainder.")
})
```

## 3. Benchmarks + accuracy

Hold out from **2008 Q1** onward. Fit mean / naïve / seasonal naïve / drift; compare RMSE.

```{r ex-bench, exercise = TRUE, exercise.lines = 18}
# train and test are available
# model MEAN, NAIVE, SNAIVE, RW(~drift()), forecast, accuracy, arrange by RMSE

```

```{r ex-bench-solution}
fit <- train |>
  model(
    mean = MEAN(Beer),
    naive = NAIVE(Beer),
    snaive = SNAIVE(Beer),
    drift = RW(Beer ~ drift())
  )
fc <- fit |> forecast(h = nrow(test))
accuracy(fc, beer) |> arrange(RMSE)
```

```{r ex-bench-check}
grade_this({
  pass_if(~ is.data.frame(.result) || inherits(.result, "tbl_df"),
          message = "Good — inspect which method has the lowest RMSE.")
  pass("Run the accuracy table and note the winner. Then plot forecasts and snaive residuals.")
})
```

Optional plots (solution code):

```{r bench-plots, exercise = TRUE, exercise.lines = 10}
fit <- train |>
  model(
    mean = MEAN(Beer),
    naive = NAIVE(Beer),
    snaive = SNAIVE(Beer),
    drift = RW(Beer ~ drift())
  )
fc <- fit |> forecast(h = nrow(test))
# autoplot forecasts; gg_tsresiduals for snaive

```

```{r bench-plots-solution}
fit <- train |>
  model(
    mean = MEAN(Beer),
    naive = NAIVE(Beer),
    snaive = SNAIVE(Beer),
    drift = RW(Beer ~ drift())
  )
fc <- fit |> forecast(h = nrow(test))
fc |> autoplot(beer, level = NULL)
fit |> select(snaive) |> gg_tsresiduals()
```

```{r bench-plots-check}
grade_this({
  pass("Nice — compare forecast paths and check whether snaive residuals look like white noise.")
})
```

```{r q-snaive}
quiz(
  question(
    "For quarterly beer with clear seasonality, which benchmark is usually the strongest starting point?",
    answer("MEAN (global average)"),
    answer("NAIVE (last value)"),
    answer("SNAIVE (same season last year)", correct = TRUE),
    answer("A random shuffle of the training set"),
    allow_retry = TRUE
  )
)
```

## 4. SES hand-check (bridge to Session 1)

SES level update: $\ell_t = \alpha y_t + (1-\alpha)\ell_{t-1}$.

Given $y = (10, 12, 11)$, $\alpha = 0.5$, $\ell_0 = 10$, compute $\ell_1, \ell_2, \ell_3$.

```{r ex-ses, exercise = TRUE, exercise.lines = 12}
y <- c(10, 12, 11)
alpha <- 0.5
ell0 <- 10
# fill ell; expect c(10, 11, 11)

```

```{r ex-ses-solution}
y <- c(10, 12, 11)
alpha <- 0.5
ell0 <- 10
ell <- numeric(3)
ell[1] <- alpha * y[1] + (1 - alpha) * ell0
for (t in 2:3) ell[t] <- alpha * y[t] + (1 - alpha) * ell[t - 1]
ell
```

```{r ex-ses-check}
grade_this({
  pass_if_equal(c(10, 11, 11), message = "Correct: levels are 10, 11, 11.")
  fail_if(~ !is.numeric(.result) || length(.result) != 3,
          message = "Return a length-3 numeric vector `ell`.")
  fail("Check the recursion: start from ell0 = 10 with alpha = 0.5.")
})
```

```{r q-ses}
quiz(
  question(
    "SES point forecasts from the last level $\\ell_T$ are:",
    answer("Flat: $\\hat{y}_{T+h|T} = \\ell_T$ for all $h \\ge 1$", correct = TRUE),
    answer("Always linearly trending"),
    answer("Always equal to the sample mean"),
    answer("Seasonally repeating the last year only"),
    allow_retry = TRUE
  )
)
```

## 5. Optional — multiplicative season?

```{r gas-plot}
aus_production |>
  filter(!is.na(Gas)) |>
  autoplot(Gas) +
  labs(title = "Gas production")
```

```{r q-gas}
quiz(
  question(
    "Looking at Gas, seasonal swings that grow with the level suggest:",
    answer("Additive seasonality only"),
    answer("Multiplicative seasonality (or a variance-stabilizing transform) is more plausible", correct = TRUE),
    answer("No seasonality at all"),
    answer("That MAE equals RMSE by definition"),
    allow_retry = TRUE
  )
)
```

## Your turn

1. Pick another series in `tsibbledata` and repeat the graphics checklist (new exercise box below).
2. Which benchmark would you defend for that series, and why?
3. List two TSA gaps you want Session 1–2 to reinforce.

```{r ex-your-turn, exercise = TRUE, exercise.lines = 14}
# Example idea: aus_production Gas, or another tsibble series
# autoplot / gg_season / gg_subseries / ACF

```

```{r ex-your-turn-solution}
gas <- aus_production |>
  filter(!is.na(Gas)) |>
  select(Quarter, Gas)

gas |> autoplot(Gas)
gas |> gg_season(Gas)
gas |> gg_subseries(Gas)
gas |> ACF(Gas) |> autoplot()
```

```{r ex-your-turn-check}
grade_this({
  pass("Good exploration. Bring your 3 bullets + preferred benchmark to class.")
})
```

## Done

Submit a **`.txt` completion file** to Blackboard Ultra — **not** HTML.

The RStudio Viewer often breaks Shiny downloads (you may see `dl_lab00.html` / "Site wasn't available").  
**Use the Save button below** — it writes a real `.txt` to your Downloads folder.

1. Enter your details.
2. Click **Save .txt to Downloads**.
3. Upload that `Lab00_TSA_Review_....txt` file to Blackboard.

Do **not** upload `00-tsa-review-learnr.html` or `dl_lab00.html`.

```{r receipt-ui, context="render"}
shiny::tagList(
  shiny::textInput("stu_name", "Full name", placeholder = "Given name + surname"),
  shiny::textInput("stu_id", "Student ID or IE email", placeholder = "xxx@student.ie.edu"),
  shiny::textInput("stu_section", "Section / group (optional)", placeholder = "PPLEDBA"),
  shiny::br(),
  shiny::actionButton(
    "write_lab00",
    "Save .txt to Downloads",
    class = "btn-primary btn-lg"
  ),
  shiny::br(), shiny::br(),
  shiny::verbatimTextOutput("write_lab00_msg"),
  shiny::tags$hr(),
  shiny::tags$p(shiny::tags$strong("Fallback:"), " copy the text below into a file named ",
                shiny::tags$code("Lab00_TSA_Review_YourName.txt"), " and upload that."),
  shiny::verbatimTextOutput("receipt_preview")
)
```

```{r receipt-server, context="server"}
`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0 || !nzchar(as.character(x)[1])) y else x
}

expected_steps <- c(
  "ex-graphics", "ex-stl", "ex-bench", "bench-plots",
  "ex-ses", "ex-your-turn",
  "q-acf", "q-snaive", "q-ses", "q-gas"
)

build_receipt_txt <- function(name, stu_id, stu_section) {
  state <- shiny::isolate(
    shiny::reactiveValuesToList(learnr::get_tutorial_state())
  )
  info <- tryCatch(
    learnr::get_tutorial_info(),
    error = function(e) list(tutorial_id = "00-tsa-review-learnr")
  )

  done <- intersect(expected_steps, names(state))
  missing <- setdiff(expected_steps, names(state))

  lines_done <- if (length(done)) {
    vapply(sort(done), function(lab) {
      item <- state[[lab]]
      sprintf(
        "- %s [%s] submitted @ %s",
        lab,
        item$type %||% "?",
        item$timestamp %||% ""
      )
    }, character(1))
  } else {
    "- (no submissions recorded yet — finish exercises/quizzes, then Save again)"
  }

  lines_missing <- if (length(missing)) paste0("- ", missing) else "- (none)"
  stamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")

  c(
    "Lab 0 — TSA review (learnr) — COMPLETION EVIDENCE",
    "Course: Forecasting for Time Series (PPLEDBA), Fall 2026",
    sprintf("Tutorial: %s", info$tutorial_id %||% "00-tsa-review-learnr"),
    "FILE_TYPE: plain text (.txt) — do not submit HTML",
    "Purpose: proof of completion for Blackboard Ultra",
    "SUBMIT_TO: Blackboard Ultra — Lab 0 assignment",
    "",
    sprintf("Full name: %s", name),
    sprintf("Student ID / email: %s", stu_id),
    sprintf("Section / group: %s", stu_section),
    sprintf("Exported at: %s", stamp),
    sprintf("R version: %s", paste(R.version$major, R.version$minor, sep = ".")),
    "",
    sprintf("COMPLETED_STEPS: %d / %d", length(done), length(expected_steps)),
    "",
    "Submitted in this session:",
    lines_done,
    "",
    "Not yet submitted:",
    lines_missing,
    "",
    "I confirm I worked through Lab 0 (interactive learnr) and upload this .txt as completion evidence."
  )
}

receipt_filename <- function(name) {
  nm <- gsub("[^A-Za-z0-9._-]+", "_", trimws(name %||% "student"))
  if (!nzchar(nm)) nm <- "student"
  paste0("Lab00_TSA_Review_", nm, ".txt")
}

receipt_dir <- function() {
  candidates <- c(
    path.expand("~/Downloads"),
    path.expand("~/Desktop"),
    path.expand("~")
  )
  for (d in candidates) {
    if (dir.exists(d) && file.access(d, 2) == 0) return(d)
  }
  getwd()
}

write_msg <- shiny::reactiveVal("Enter your name, then click Save .txt to Downloads.")
preview_txt <- shiny::reactiveVal("(Save once to refresh this preview, or fill your name and click Save.)")

output$write_lab00_msg <- shiny::renderText(write_msg())
output$receipt_preview <- shiny::renderText(preview_txt())

shiny::observeEvent(input$write_lab00, {
  name <- trimws(input$stu_name %||% "")
  if (!nzchar(name)) {
    write_msg("Enter your full name first.")
    return()
  }

  txt <- build_receipt_txt(
    name,
    trimws(input$stu_id %||% ""),
    trimws(input$stu_section %||% "")
  )
  fname <- receipt_filename(name)
  path <- file.path(receipt_dir(), fname)
  writeLines(txt, path, useBytes = FALSE)

  preview_txt(paste(txt, collapse = "\n"))
  write_msg(paste0(
    "Saved plain-text evidence (.txt):\n",
    normalizePath(path, winslash = "/", mustWork = FALSE),
    "\n\nUpload THIS .txt to Blackboard. Ignore any dl_lab00.html download."
  ))
})
```

Companion static lab: [`00-tsa-review.Rmd`](00-tsa-review.Rmd).
