---
title: "Session 4 — Non-seasonal ARIMA (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.  
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

ARIMA(1,1,0): $\nabla y_t = 0.2 + 0.5\,\nabla y_{t-1}$.  
Levels $y = (10, 12, 11, 14, 13)$.

```{r}
y <- c(10, 12, 11, 14, 13)
dy <- diff(y)                 # length 4: t = 2..5
phi <- 0.5
c0 <- 0.2
# Predictions of the difference use the *previous* difference:
# for dy[t], use dy[t-1] (skip the first difference — no lag yet).
pred_dy <- c(NA_real_, c0 + phi * dy[-length(dy)])
tibble::tibble(
  t = 2:5,
  dy = dy,
  pred_dy = pred_dy,
  y_prev = y[-length(y)],
  yhat = y[-length(y)] + pred_dy,
  y = y[-1]
)
```

Check $t=5$: $\widehat{\nabla y}_5 = 0.2 + 0.5\times 3 = 1.7$, so $\hat y_5 = 14 + 1.7 = 15.7$.

# Spain GDP

```{r}
# Keep a single-series tsibble (Year index only). Leaving `Country` as a key
# breaks some residual helpers on common Posit Cloud package stacks.
spain <- global_economy |>
  filter(Country == "Spain") |>
  select(Year, GDP)

spain |>
  autoplot(GDP) +
  scale_y_log10() +
  labs(title = "Spain GDP (log scale)", y = "GDP")

spain |>
  mutate(g = difference(log(GDP))) |>
  filter(!is.na(g)) |>
  autoplot(g) +
  labs(title = "Growth rates: diff log(GDP)", y = NULL)
```

# Train / test split (origin 2010)

```{r}
train <- spain |> filter(Year <= 2010)
test <- spain |> filter(Year > 2010)

fit <- train |>
  model(
    rw = RW(log(GDP) ~ drift()),
    arima_auto = ARIMA(log(GDP) ~ PDQ(0, 0, 0)),
    arima_110 = ARIMA(log(GDP) ~ pdq(1, 1, 0) + PDQ(0, 0, 0))
  )

glance(fit) |> arrange(AICc)
fc <- fit |> forecast(h = nrow(test))
accuracy(fc, spain) |> arrange(RMSE)
fc |>
  autoplot(spain, level = NULL) +
  labs(title = "Spain GDP forecasts (trained <= 2010)")
```

# Residuals

`residuals()` on a transformed ARIMA can return an empty series on some Posit Cloud package versions. Build innovations from `augment()` instead (`.innov` when present, otherwise a log residual).

```{r}
aug <- as_tibble(augment(fit))
aug <- aug[aug$.model == "arima_auto", , drop = FALSE]

if (".innov" %in% names(aug) && any(!is.na(aug$.innov))) {
  e <- aug$.innov
} else if (any(!is.na(aug$.fitted)) && any(aug$GDP > 0, na.rm = TRUE)) {
  # .fitted may be on the GDP scale or already on the log scale
  fitted_big <- stats::median(abs(aug$.fitted), na.rm = TRUE) > 1000
  e <- if (fitted_big) log(aug$GDP) - log(aug$.fitted) else log(aug$GDP) - aug$.fitted
} else {
  e <- aug$.resid
}

innov <- tibble::tibble(Year = aug$Year, e = e)
innov <- innov[!is.na(innov$e), ]
innov <- as_tsibble(innov, index = Year)

autoplot(innov, e) + labs(title = "ARIMA auto — innovation residuals", y = NULL)
innov |> ACF(e) |> autoplot() + labs(title = "ACF of innovations")
```

# Your turn

1. Repeat for another country in `global_economy` (filter, then `select(Year, GDP)`).
2. Force $d=0$ vs $d=1$ and compare residual ACF + test RMSE.
3. Write 3 sentences recommending a model to a policy analyst.
