---
title: "Session 4 — Non-seasonal ARIMA (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.  
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

ARIMA(1,1,0): $\nabla y_t = 0.2 + 0.5\,\nabla y_{t-1}$.  
Levels $y = (10, 12, 11, 14, 13)$.

```{r}
y <- c(10, 12, 11, 14, 13)
dy <- diff(y)                 # length 4: t = 2..5
phi <- 0.5
c0 <- 0.2
# Predictions of the difference use the *previous* difference:
# for dy[t], use dy[t-1] (skip the first difference — no lag yet).
pred_dy <- c(NA_real_, c0 + phi * dy[-length(dy)])
tibble::tibble(
  t = 2:5,
  dy = dy,
  pred_dy = pred_dy,
  y_prev = y[-length(y)],
  yhat = y[-length(y)] + pred_dy,
  y = y[-1]
)
```

Check $t=5$: $\widehat{\nabla y}_5 = 0.2 + 0.5\times 3 = 1.7$, so $\hat y_5 = 14 + 1.7 = 15.7$.

# Spain GDP

```{r}
# Keep a single-series tsibble (Year index only). Leaving `Country` as a key
# breaks some residual helpers on common Posit Cloud package stacks.
spain <- global_economy |>
  filter(Country == "Spain") |>
  select(Year, GDP)

spain |>
  autoplot(GDP) +
  scale_y_log10() +
  labs(title = "Spain GDP (log scale)", y = "GDP")

spain |>
  mutate(g = difference(log(GDP))) |>
  filter(!is.na(g)) |>
  autoplot(g) +
  labs(title = "Growth rates: diff log(GDP)", y = NULL)
```

# Train / test split (origin 2010)

```{r}
train <- spain |> filter(Year <= 2010)
test <- spain |> filter(Year > 2010)

fit <- train |>
  model(
    rw = RW(log(GDP) ~ drift()),
    arima_auto = ARIMA(log(GDP) ~ PDQ(0, 0, 0)),
    arima_110 = ARIMA(log(GDP) ~ pdq(1, 1, 0) + PDQ(0, 0, 0))
  )

glance(fit) |> arrange(AICc)
fc <- fit |> forecast(h = nrow(test))
accuracy(fc, spain) |> arrange(RMSE)
fc |>
  autoplot(spain, level = NULL) +
  labs(title = "Spain GDP forecasts (trained <= 2010)")
```

# Residuals

`residuals()` / `augment()` can yield an empty series for `ARIMA(log(GDP))` on some Posit Cloud stacks. Read the innovation residuals stored on the fitted model instead, and plot with ggplot (not `feasts::ACF`).

```{r}
m <- fit[["arima_auto"]][[1]]
est <- m$fit$est
if (!is.null(est) && ".resid" %in% names(est)) {
  e <- est$.resid
} else {
  e <- as.numeric(stats::residuals(m$fit$model))
}
year <- tail(train$Year, length(e))
innov <- tibble::tibble(Year = year, e = e)
innov <- innov[is.finite(innov$e), ]

ggplot(innov, aes(Year, e)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line() +
  labs(title = "ARIMA auto — innovation residuals", y = NULL)

acf_obj <- stats::acf(innov$e, plot = FALSE, na.action = na.pass)
acf_df <- tibble::tibble(
  lag = as.numeric(acf_obj$lag),
  acf = as.numeric(acf_obj$acf)
)
acf_df <- acf_df[acf_df$lag > 0, ]
band <- 2 / sqrt(nrow(innov))
ggplot(acf_df, aes(lag, acf)) +
  geom_hline(yintercept = c(-band, 0, band), linetype = c(2, 1, 2), colour = "steelblue") +
  geom_col(width = 0.25) +
  labs(title = "ACF of innovations", x = "lag", y = "acf")
```

# Your turn

1. Repeat for another country in `global_economy` (filter, then `select(Year, GDP)`).
2. Force $d=0$ vs $d=1$ and compare residual ACF + test RMSE.
3. Write 3 sentences recommending a model to a policy analyst.
