---
title: "Session 4 — Non-seasonal ARIMA (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Lecture slides emphasise theory, figures, and hand calculations.  
> This lab is where you **type and run** the tidy `fable` workflow (and check the toy arithmetic).

# Packages

```{r}
library(fpp3)
```

# Hand calculations (match the slides)

ARIMA(1,1,0): $\nabla y_t = 0.2 + 0.5\,\nabla y_{t-1}$.  
Levels $y = (10, 12, 11, 14, 13)$.

```{r}
y <- c(10, 12, 11, 14, 13)
dy <- diff(y)
phi <- 0.5
c0 <- 0.2
# one-step predictions of differences for t = 3,4,5 (indices 2,3,4 of dy)
pred_dy <- c0 + phi * dplyr::lag(dy)
level_fcast <- y[-length(y)] + pred_dy
tibble::tibble(
  t = 2:5,
  dy = dy,
  pred_dy = pred_dy,
  y_prev = y[-length(y)],
  yhat = level_fcast,
  y = y[-1]
)
```

Check $t=5$: $\widehat{\nabla y}_5 = 1.7$, $\hat y_5 = 15.7$.

# Spain GDP

```{r}
spain <- global_economy |> filter(Country == "Spain")
spain |> autoplot(GDP) + scale_y_log10() +
  labs(title = "Spain GDP (log scale)", y = "GDP")

spain |>
  mutate(g = difference(log(GDP))) |>
  autoplot(g) +
  labs(title = "Growth rates: diff log(GDP)", y = NULL)
```

# Train / test split (origin 2010)

```{r}
train <- spain |> filter(Year <= 2010)
test <- spain |> filter(Year > 2010)

fit <- train |>
  model(
    rw = RW(log(GDP) ~ drift()),
    arima_auto = ARIMA(log(GDP) ~ PDQ(0, 0, 0)),
    arima_110 = ARIMA(log(GDP) ~ pdq(1, 1, 0) + PDQ(0, 0, 0))
  )

glance(fit) |> arrange(AICc)
fc <- fit |> forecast(h = nrow(test))
# Evaluate against the full series (fable convention used on the slides)
accuracy(fc, spain) |> arrange(RMSE)
fc |> autoplot(spain, level = NULL) +
  labs(title = "Spain GDP forecasts (trained <= 2010)")
```

# Residuals

```{r}
fit |> select(arima_auto) |> gg_tsresiduals()
```

# Your turn

1. Repeat for another country in `global_economy`.
2. Force $d=0$ vs $d=1$ and compare residual ACF + test RMSE.
3. Write 3 sentences recommending a model to a policy analyst.
