---
title: "Session 13 - Prophet vs ARIMA (R lab)"
subtitle: "Forecasting for Time Series (PPLEDBA)"
author: "Prof. Dae-Jin Lee (daelee@faculty.ie.edu)"
date: "Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
```

> **Slides vs lab.** Slides expand Prophet theory (trend, Fourier season, holidays).
> This lab runs the Beer bake-off. **Prophet is optional**: the guarded chunk skips
> cleanly if `fable.prophet` is not installed.

# Packages

```{r}
library(fpp3)
```

# Recall: Prophet model

\[
y_t = g(t) + s(t) + h(t) + \varepsilon_t
\]

- $g(t)$: piecewise-linear trend with changepoints  
- $s(t)$: Fourier seasonality (`season(period, order)`)  
- $h(t)$: holiday dummies (optional)  

For quarterly Beer, set `period = 4` explicitly (defaults assume daily-or-finer data).

# Beer bake-off: SNAIVE, ETS, ARIMA, DHR

Same hold-out as Sessions 5 / 12: train through **2007 Q4**.

```{r}
beer <- aus_production |> filter(!is.na(Beer))
train <- beer |> filter_index(. ~ "2007 Q4")
test <- beer |> filter_index("2008 Q1" ~ .)

fit <- train |>
  model(
    snaive = SNAIVE(Beer),
    ets = ETS(Beer),
    arima = ARIMA(Beer),
    dhr = ARIMA(Beer ~ fourier(K = 2) + PDQ(0, 0, 0))
  )
fc <- fit |> forecast(h = nrow(test))
accuracy(fc, beer) |> select(.model, RMSE, MAE, MAPE) |> arrange(RMSE)
fc |> autoplot(beer, level = NULL) +
  labs(title = "Method bake-off on Beer (hold-out from 2008 Q1)")
```

# Optional: Prophet (guarded)

```{r}
if (requireNamespace("fable.prophet", quietly = TRUE)) {
  library(fable.prophet)
  fit_prophet <- train |>
    model(
      prophet = prophet(Beer ~ season(period = 4, order = 2))
    )
  fc_prophet <- fit_prophet |> forecast(h = nrow(test))
  accuracy(fc_prophet, beer) |> select(.model, RMSE, MAE, MAPE)
  # Components (trend / season) - useful for communication slides
  fit_prophet |> components() |> autoplot()
} else {
  message("fable.prophet not installed - skipping Prophet chunk")
  message("Install with install.packages('fable.prophet') if you have a Stan toolchain.")
}
```

# Your turn

1. Fill a small table: method, RMSE, interpretability (H/M/L), one residual note.
2. If Prophet ran: does it beat ETS/ARIMA on this hold-out? Relate to FPP3's warning.
3. When would you present Prophet components to a product manager? When would you insist on an ARIMA+regression bake-off for a risk audience?
